---
title: "Untitled"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction

  Urban Ministries is a non-profit organization that seeks to provide food, job opportunities, and other services to the homeless community of Durham and the surrounding area. Using the data provided by UM, I hope to provide meaningful insight into how the services they provide are affected by various social logical factors, such as age, race, and gender.


```{r, echo=FALSE}
#Load in necessary packages
library(tidyverse)
library(ggplot2)
library(data.table)
```

```{r}
#load in dataset
client.data=fread(file="C:/Users/alex/Documents/GitHub/datasci611/data/project2_2019/CLIENT_191102.tsv") %>% select(`Client Unique ID`, `Client Age at Entry`, `Client Gender`, `Client Primary Race`, `Client Ethnicity`)
disability.data=fread(file="C:/Users/alex/Documents/GitHub/datasci611/data/project2_2019/DISABILITY_ENTRY_191102.tsv")
analysis.data = disability.data %>% left_join(client.data, by = "Client Unique ID") %>% distinct() %>% mutate(counts = as.numeric(as.factor(`Disability Type (Entry)`)))
for(i in 1:nrow(analysis.data)){
if(analysis.data[i,5] == "No (HUD)"){
  analysis.data[i,16] = 0
}}
```

```{r}
ggplot(analysis.data, aes(x=counts)) + geom_bar(stat = "bin")
```

It would appear that overall, the most common disability among UMD's clients are mental health problems.